Dynamic Modeling of Cell-Free Biochemical Networks using Effective Kinetic Models

### 2015-02-16 revision_v3/Paper_v3.tex
Submission of revised v2

Mathematical modeling has long contributed to our understanding of metabolism. Decades before the genomics revolution, mechanistically, structured metabolic models arose from the desire to predict microbial phenotypes resulting from changes in intracellular or extracellular states \citep{1976_fredrickson_BiotechBioeng}. 
The single cell \textit{E. coli} models of Shuler and coworkers pioneered the construction of large-scale, dynamic metabolic models that incorporated multiple, regulated catabolic and anabolic pathways constrained by experimentally determined kinetic parameters \citep{1984_domach_shuler_BiotechBioeng_01}. 
Shuler and coworkers generated many single cell kinetic models, including single cell models of eukaryotes \citep{1989_steinmeyer_shuler_ChemEngSci,1992_wu_shuler_AnnNYAcadSci}, minimal cell architectures \citep{2004_castellanos_shuler_PNAS}, as well as DNA sequence based whole-cell models of \textit{E. coli} \citep{2008_atlas_shuler_IETSysBio}.
Conversely, highly abstracted kinetic frameworks, such as the cybernetic framework, represented a paradigm shift, viewing cells as growth-optimizing strategists \citep{1985_dhurjati_ramkrishna_tsao_BiotechBioeng}. 
Cybernetic models have been highly successful at predicting metabolic choice behavior, e.g., diauxie behavior \citep{1986_kompala_ramkrishna_tsao_BiotechBioeng}, steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, as well as the cellular response to metabolic engineering modifications \citep{1999_varner_ramkrishna_MetaEng}. 
Unfortunately, traditional, fully structured cybernetic models also suffer from an identifiability challenge, as both the kinetic parameters and an abstracted model of cellular objectives must be estimated simultaneously. 
However, recent cybernetic formulations from Ramkrishna and colleagues have successfully treated this identifiability challenge through elementary mode reduction \cite{2009_song_ramkrishna_BiotechBioeng,Song:2011aa}.

@@

The proposed modeling strategy shares features with other popular techniques, but also has several key differences. 
At its core, our effective modeling approach is similar to regulatory constraint-based methods, and to the cybernetic modeling paradigm developed by Ramkrishna and colleagues.
Covert, Palsson and coworkers drastically improved the predictability of constraint-based approaches by integrating Boolean rules into the calculation of metabolic fluxes \citep{Covert:2001aa}.
If the regulated intracellular flux problem is coupled with time-dependent extracellular balances, these models can predict complex behavior such as diauxie growth or the switch between 
aerobic and anaerobic metabolism. Another important feature of this approach is that it scales with biological complexity. 
For example, Covert \emph{et al.} showed that a genome-scale model of \emph{E.~coli} augmented with a Boolean rule layer, 
correctly predicted approximately 80\% of the outcomes of a high-throughput growth phenotyping experiment in \emph{E. coli}.
Further, they showed that they could learn new biology by iteratively refining the model and its associated rules \citep{Covert:2004aa}. 
However, while regulated flux balance analysis is a powerful technique, it does not easily allow the calculation of time-resolved metabolite abundance. 
Additionally, the Boolean rules which populate the regulatory layer are limited to ON/OFF decisions; for qualitative predictions of gene expression this is a reasonable limitation. 
However, Boolean rules will likely be less effective at capturing dynamic allosteric regulation in a cell-free metabolic system. 
On the other hand, the strength of cybernetic models is the integration of optimal metabolic control strategies with traditional kinetic pathway modeling. 
Cybernetic models are highly predictive; they have successfully predicted mutant behavior from limited wild-type data \citep{1999_varner_ramkrishna_MetaEng,Varner:2000aa,Song:2012aa}, steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, strain specific metabolic function \citep{Song:2011aa} and have been used in bioprocess control applications \citep{Gadkar:2003aa}.
However, cybernetic control strategies are not mechanistic, instead they are the output of an optimal decision with respect to a set of hypothetical physiological objectives. 
Thus, they are abstractions which are difficult to translate into a specific biological mechanism. 
Our approach addresses the shortcomings of both regulatory constraint-based models and cybernetic models. 
First, similar to cybernetic models, the core of our approach is a kinetic model. Thus, we are able to directly calculate the time evolution of metabolism, for example the dynamic abundance of network metabolites. Second, similar to regulatory flux balance analysis, our control laws describe specific mechanistic motifs, such as activation or inhibition of enzyme activity. 
However, our rules are continuous, thus they potentially allow a finer grained description of metabolic regulation and control mechanisms.
Lastly, we can naturally incorporate unmodeled factors and categorical factors or combinations thereof into our control law formulations. 
Though requiring a more complex description of cellular metabolism, our approach may even be extended to simulate cell-based systems by incorporating the same control laws into transcription factor activation and gene expression regulation.

The proposed modeling framework also differs appreciably from previously established kinetic approximations of complex biochemical network behavior. 
Such frameworks replace parameter dense mechanistic kinetic expressions with heuristics quantifying the relationship between metabolic rate and metabolite effectors. 
A review of approximative kinetic formats can be found in \cite{2005_heijnen_BiotechBioeng}.  
These approaches arose in response to uncertainties associated with obtaining correct mechanistic kinetic expressions and parameters of \textit{in vivo} systems. 
Similarly, available kinetic parameters measured \textit{in vitro} may differ in a specialized cell-free \textit{in vitro} environments. 
Factors affecting kinetics, such as enzyme channeling, macromolecular crowding, and pH, are likely dramatically different in cell-free environments than in both \textit{in vivo} systems as well as typical \textit{in vitro} conditions used for parameter measurements. 
Thus, a more generalized, approximate biochemical reaction network formulation may be desirable in the case of cell-free systems. 
Our approach is similar to generalized mass action-based power law formulations of Savageau and colleagues \cite{1991_savageau_JTheorBio} and linlog kinetics of Visser \textit{et al.} \cite{2003_visser_heijnen_MetabEng} in that metabolic rates are proportional to corresponding enzyme levels modified by metabolite effectors. 
Power law and linlog approaches suffer from several limitations \cite{2005_heijnen_BiotechBioeng}. 
Power law reaction rates do not capture saturation effects and become infinite for small concentrations of inhibitory regulators. 
Linlog kinetics also become ill-defined when effector concentrations go to zero. 
Also, models employing linlog kinetics typically rely on an experimentally determined reference state to describe dynamics taking place after a perturbation to a steady-state. 
Our framework does not suffer from such drawbacks.  
Moreover, cell-free systems are unlikely to satisfy such a steady-state approximation after extract preparation and prior to culture initiation. 
Our framework is similar to the generic kinetic formulation from Hadlich \textit{et al.} \cite{2009_hadlich_wiechert_MetabEng}, 
but differs in its inclusion of cooperative effects as well as proposes a simplified integration of competition amongst allosteric effectors using max/min rules. 
In summary, our proposed framework offers an effective kinetic approximation that captures saturation effects and allosteric competition within cell-free systems that may also be extensible to \textit{in vivo} metabolic and gene regulatory networks.

There are several critical questions that should be explored following this proof-of-concept study.
It is unclear how parameter identification will scale to genome-scale networks, and second it is unclear how we will identify allosteric connectivity at a genome-scale. 
The enzymatic connectivity for genome-scale cell-free networks can easily be established by stripping away the growth and cell wall machinery from whole cell genome reconstructions.
Then metabolic fluxes can be transformed into kinetic expressions using heuristics such multiple saturation kinetics, which are then modified by our rule-based control variables.  
This leaves a large number of unknown kinetic constants that must be estimated from time-resolved metabolite measurements. 
Ensemble modeling is a well-established approach for parameter identification in large-scale deterministic models. 
Liao and coworkers developed a method that generates an ensemble of kinetic models that all approach the same steady-state, one determined by fluxomics measurements \cite{Tran:2008aa}. 
The best subpopulation of candidate models were selected based on their agreement with further measurements of genetically perturbed systems. 
Our work relies on heuristic search optimization to identify kinetic models consistent with steady-state and dynamic time-series measurements of cellular species \citep{Luan:2007aa,Song:2009aa,Tasseff:2010aa,Tasseff:2011aa,Nayak:2011aa,Lequieu:2011aa}. 
Instead of estimating a single yet highly uncertain parameter set, both approaches estimate an ensemble of parameter sets whose model behavior recapitulates experimental measurements. 
Here, we showed that particle swarm optimization quickly identified an ensemble of model parameters, at least for proof-of-concept metabolic networks using synthetic data. 
This suggested that we can expect reasonable model predictions, despite only partial parameter knowledge, as network size grows if we have properly designed experiments. 
Though we expect computational complexity will scale poorly with network size, we are optimistic that large-scale, predictive models of metabolism are possible. There is evidence to suggest that achieving a quantitative understanding of complex biological systems should not require complete parametric knowledge.
Brown and Sethna showed in a model of signal transduction that good predictions were possible despite only order of magnitude estimates of parameter values \citep{Brown:2003aa}.
Sethna and coworkers later showed that model performance is often controlled by only a few parameter combinations, a characteristic seemingly universal to multi-parameter 
models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
We have also demonstrated \textit{sloppy} behavior in a wide variety of signal transduction processes \citep{Luan:2007aa,Song:2009aa,Tasseff:2010aa,Tasseff:2011aa,Nayak:2011aa,Lequieu:2011aa}.
Thus, given our previous experience with models containing hundreds of unknown parameters, we expect parameter estimation to be a manageable challenge assuming we have good quality experimental data.

@@

We modified our PSO implementation to simultaneously search over kinetic parameters and putative model control structures.
In the combined case, each particle potentially carried a different model realization in addition to a different kinetic parameter vector. 
We kept the update rules the same (along with the update parameters). 
Thus, each particle competed on the basis of goodness of fit, which allowed different model structures to contribute to the overall behavior of the swarm. 
We considered five possible model structures (A through E), where network A was the correct formulation (used to generate the synthetic data). 
We considered a population of 100 particles, where each particle in the swarm was assigned a model structure, and a random parameter vector. 
The optimization simulations shown in Figure \ref{fig-control-search} required several hours to complete on a single CPU Apple workstation (Apple, Cupertino, CA; OS X v10.10).  
The PSO algorithm, model equations, and the objective function were encoded and solved in the Octave programming language (v 3.8.1; www.octave.org).

>>>>

Mathematical modeling has long contributed to our understanding of metabolism. Decades before the genomics revolution, mechanistically, structured metabolic models arose from the desire to predict microbial phenotypes resulting from changes in intracellular or extracellular states \citep{1976_fredrickson_BiotechBioeng}. 
The single cell \textit{E. coli} models of Shuler and coworkers pioneered the construction of large-scale, dynamic metabolic models that incorporated multiple, regulated catabolic and anabolic pathways constrained by experimentally determined kinetic parameters \citep{1984_domach_shuler_BiotechBioeng_01}. 
Shuler and coworkers generated many single cell kinetic models, including single cell models of eukaryotes \citep{1989_steinmeyer_shuler_ChemEngSci,1992_wu_shuler_AnnNYAcadSci}, minimal cell architectures \citep{2004_castellanos_shuler_PNAS}, as well as DNA sequence based whole-cell models of \textit{E. coli} \citep{2008_atlas_shuler_IETSysBio}.
Conversely, highly abstracted kinetic frameworks, such as the cybernetic framework, represented a paradigm shift, viewing cells as growth-optimizing strategists \citep{1985_dhurjati_ramkrishna_tsao_BiotechBioeng}. 
Cybernetic models have been highly successful at predicting metabolic choice behavior, e.g., diauxie behavior \citep{1986_kompala_ramkrishna_tsao_BiotechBioeng}, steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, as well as the cellular response to metabolic engineering modifications \citep{1999_varner_ramkrishna_MetaEng}. 
Unfortunately, traditional, fully structured cybernetic models also suffer from an identifiability challenge, as both the kinetic parameters and an abstracted model of cellular objectives must be estimated simultaneously. 
However, recent cybernetic formulations from Ramkrishna and colleagues have successfully treated this identifiability challenge through elementary mode reduction, though the techniques replace detailed biological mechanism with an optimization heuristic \cite{2009_song_ramkrishna_BiotechBioeng,Song:2011aa}.

@@

The proposed modeling strategy shares features with other popular techniques, but also has several key differences. 
At its core, our effective modeling approach is similar to regulatory constraint-based methods, and to the cybernetic modeling paradigm developed by Ramkrishna and colleagues.
Covert, Palsson and coworkers drastically improved the predictability of constraint-based approaches by integrating Boolean rules into the calculation of metabolic fluxes \citep{Covert:2001aa}.
If the regulated intracellular flux problem is coupled with time-dependent extracellular balances, these models can predict complex behavior such as diauxie growth or the switch between 
aerobic and anaerobic metabolism. Another important feature of this approach is that it scales with biological complexity. 
For example, Covert \emph{et al.} showed that a genome-scale model of \emph{E.~coli} augmented with a Boolean rule layer, 
correctly predicted approximately 80\% of the outcomes of a high-throughput growth phenotyping experiment in \emph{E. coli}.
Further, they showed that they could learn new biology by iteratively refining the model and its associated rules \citep{Covert:2004aa}. 
However, while regulated flux balance analysis is a powerful technique, it does not easily allow the calculation of time-resolved metabolite abundance. 
Additionally, the Boolean rules which populate the regulatory layer are limited to ON/OFF decisions; for qualitative predictions of gene expression this is a reasonable limitation. 
However, Boolean rules will likely be less effective at capturing dynamic allosteric regulation in a cell-free metabolic system. 
On the other hand, the strength of cybernetic models is the integration of optimal metabolic control heuristics with traditional kinetic pathway modeling. 
Cybernetic models are highly predictive; they have successfully predicted mutant behavior from limited wild-type data \citep{1999_varner_ramkrishna_MetaEng,Varner:2000aa,Song:2012aa}, 
steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, strain specific metabolic function \citep{Song:2011aa} and have been used in bioprocess control applications \citep{Gadkar:2003aa}.
However, cybernetic control heuristics are not mechanistic, instead they are the output of an optimal decision with respect to a set of hypothetical physiological objectives. 
Thus, they are abstractions which are difficult to translate into a specific biological mechanism. 
Our approach addresses the shortcomings of both regulatory constraint-based models and cybernetic models. 
First, similar to cybernetic models, the core of our approach is a kinetic model. Thus, we are able to directly calculate the time evolution of metabolism, for example the dynamic abundance of network metabolites. Second, similar to regulatory flux balance analysis, our control laws describe specific mechanistic motifs, such as activation or inhibition of enzyme activity. 
However, our rules are continuous, thus they potentially allow a finer grained description of metabolic regulation and control mechanisms.
Lastly, we can naturally incorporate unmodeled factors and categorical factors or combinations thereof into our control law formulations. 
Though requiring a more complex description of cellular metabolism, our approach may even be extended to simulate cell-based systems by incorporating the same control laws into transcription factor activation and gene expression regulation.

There are several critical questions that should be explored following this proof-of-concept study.
It is unclear how parameter identification will scale to genome-scale networks, and second it is unclear how we will identify allosteric connectivity at a genome-scale. 
The enzymatic connectivity for genome-scale cell-free networks can easily be established by stripping away the growth and cell wall machinery from whole cell genome reconstructions.
Then metabolic fluxes can be transformed into kinetic expressions using heuristics such multiple saturation kinetics, which are then modified by our rule-based control variables.  
This leaves a large number of unknown kinetic constants that must be estimated from time-resolved metabolite measurements. 
Ensemble modeling is a well-established approach for parameter identification in large-scale deterministic models. 
Liao and coworkers developed a method that generates an ensemble of kinetic models that all approach the same steady-state, one determined by fluxomics measurements \cite{Tran:2008aa}. 
The best subpopulation of candidate models were selected based on their agreement with further measurements of genetically perturbed systems. 
Our work relies on heuristic search optimization to identify kinetic models consistent with steady-state and dynamic time-series measurements of cellular species \citep{Luan:2007aa,Song:2009aa,Tasseff:2010aa,Tasseff:2011aa,Nayak:2011aa,Lequieu:2011aa}. 
Instead of estimating a single yet highly uncertain parameter set, both approaches estimate an ensemble of parameter sets whose model behavior recapitulates experimental measurements. 
Here, we showed that particle swarm optimization quickly identified an ensemble of model parameters, at least for proof-of-concept metabolic networks using synthetic data. 
This suggested that we can expect reasonable model predictions, despite only partial parameter knowledge, as network size grows if we have properly designed experiments. 
Brown and Sethna showed in a model of signal transduction that good predictions were possible despite only order of magnitude estimates of parameter values \citep{Brown:2003aa}.
Sethna and coworkers later showed that model performance is often controlled by only a few parameter combinations, 
a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
We have also demonstrated \textit{sloppy} behavior in a wide variety of signal transduction processes \citep{Luan:2007aa,Song:2009aa,Tasseff:2010aa,Tasseff:2011aa,Nayak:2011aa,Lequieu:2011aa}.
Thus, given our previous experience with models containing hundreds of unknown parameters, we expect parameter estimation to be a manageable challenge assuming 
we have good quality experimental data.

@@

We modified our PSO implementation to simultaneously search over kinetic parameters and putative model control structures.
In the combined case, each particle potentially carried a different model realization in addition to a different kinetic parameter vector. 
We kept the update rules the same (along with the update parameters). 
Thus, each particle competed on the basis of goodness of fit, which allowed different model structures to contribute to the overall behavior of the swarm. 
We considered five possible model structures (A through E), where network A was the correct formulation (used to generate the synthetic data). 
We considered a population of 100 particles, where each particle in the swarm was assigned a model structure, and a random parameter vector.
The PSO algorithm, model equations, and the objective function were encoded and solved in the Octave programming language (v 3.8.1; www.octave.org).
